---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phonfeatures

<!-- badges: start -->
<!-- badges: end -->

`phonfeatures` is an R package which allows the user to efficiently add columns with 
articulatory features to existing data frames with phonetic characters using
the function `add_features()`. 

The package provides several generic features for all regular IPA and X-SAMPA
characters as well as for most diacritics. These generic features can be checked
using the function `feature_lookup()`. The package also provides functionality
for changing the feature values of known characters with `feature_reassign()`,
and for introducing unknown characters with `feature_assign()`.

## Installation

You can install the development version of `phonfeatures` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("rpuggaardrode/phonfeatures")
```

## Usage

### Adding features to a data frame

The most common use case is when you have a data frame containing transcriptions,
and you want to convert those transcriptions into some articulatory features.
It could fx look like this:

```{r}
library(phonfeatures)
head(test_data, 10)
```

Say you want to add columns to this data frame with information about place of
articulation and laryngeal features corresponding to the stop column. You
would proceed like this:

```{r}
new_data <- add_features(data=test_data, col='Stop', feature=c('place', 'lar'))
head(new_data, 10)
```

### Checking feature values

You can use `?add_features` to see which features are generically available,
or use `feature_lookup()` to check all the features that are generically assigned
to a character:

```{r}
feature_lookup(phon='g')
```

`feature_lookup()` can also be used to check specific feature values:

```{r}
feature_lookup(phon='g', feature=c('lar', 'place'))
```

### Changing feature values

The generic features will not be suitable for all purposes. For example, if
(like me) you are interested in Danish stops, there's a good chance that you have
been using /b d g p t k/ as a shorthand for sounds that are *not* actually
voiced and voiceless, but rather voiceless and aspirated. 

The package also provides functions for creating a new lookup table used for checking
feature values. We can update the value of `lar` for /b d g/
with the function `feature_reassign()`
and save it to a new data frame `dan_lkup` like this:

```{r}
dan_lkup <- feature_reassign(sampa=c('b', 'd', 'g'), 
                             feature='lar', val='voiceless')
```

And the values of /p t k/ can then be updated by running `feature_reassign()`
again, this time specifying that we're using the lookup table `dan_lkup`:

```{r}
dan_lkup <- feature_reassign(sampa=c('p', 't', 'k'), 
                             feature='lar', val='aspirated',
                             lookup=dan_lkup)
```

We can now use `dan_lkup` to add features to our data frame, instead of using
the generic features:

```{r}
new_data <- add_features(data=test_data, col='Stop', feature=c('place', 'lar'),
                         lookup=dan_lkup)
head(new_data, 10)
```

### IPA and X-SAMPA

You may have noticed that the phonetic character argument in `feature_reassign()`
is called `sampa`. This is because this function only works with [X-SAMPA](https://en.wikipedia.org/wiki/X-SAMPA)
characters. `add_features()` and `feature_lookup()` also use X-SAMPA characters 
by default, but these functions have an option to use IPA instead, by setting
`ipa=TRUE`. Let's see what this does to the code we just ran:

```{r}
new_data <- add_features(data=test_data, col='Stop', feature=c('place', 'lar'),
                         lookup=dan_lkup, ipa=TRUE)
head(new_data, 10)
```

It makes no difference of course! /b d g p t k/ are identical in IPA and X-SAMPA.
But this is what happens when `feature_lookup()` is used with an IPA character
which *doesn't* exist as an X-SAMPA character.

```{r error=TRUE}
feature_lookup(phon='ø', feature='length')
```

The function throws a warning saying that the character is unknown. Let's try again
with `ipa=TRUE`:

```{r}
feature_lookup(phon='ø', feature='length', ipa=TRUE)
```

Say you have a data frame with IPA characters, and the sound /ø/ is always long
in the language, so you want to change the `length` value to `long`.
`feature_reassign()` only accepts X-SAMPA characters, but fear not!
The function `ipa::ipa()` is used under the hood to convert IPA characters
to X-SAMPA characters, so you can simply reassign the corresponding X-SAMPA
character (in this case `2`):

```{r}
lkup <- feature_reassign(sampa='2', feature='length', val='long')
feature_lookup(phon='ø', feature='length', lookup=lkup, ipa=TRUE)
```

### Adding new characters

You may want to use `add_features()` with a data frame where, for one reason
or another, you have been using non-standard symbols. For example, you've been 
using `ph` instead of the standard X-SAMPA `p_h` to indicate an aspirated bilabial
stop. In this case, `feature_lookup()` will shoot you an error:

```{r error=TRUE}
feature_lookup(phon='ph')
```

Again, fear not! The `feature_assign` function can help with creating
a customized lookup table with new characters. 
This feature also allows you to copy the features from a known character, or 
a character with known diacritics, using the `copy` argument. For our `ph` case,
you would do the following:

```{r}
lkup <- feature_assign(new='ph', copy='p_h')
feature_lookup(phon='ph', lookup=lkup)
```

`feature_assign` also allows you to build your own feature sets from scratch:

```{r}
lkup <- feature_assign(new='kh',
                       feature=c('place', 'manner', 'lar'),
                       val=c('velar', 'stop', 'aspirated'))
feature_lookup(phon='kh', lookup=lkup)
```

Or you can copy a known character and change some of the features (say, if 
you've been using `th` to refer to a long, dental aspirated stop):

```{r}
lkup <- feature_assign(new='th',
                       feature=c('place', 'length'),
                       val=c('dental', 'long'),
                       copy='t_h')
feature_lookup(phon='th', lookup=lkup)
```

You can bulk add new characters using `feature_assign` assuming they can all
be copied from known characters, or that the same feature needs to be manipulated
for all of them. In the following code block, features are bulk added from 
known characters:

```{r}
lkup <- feature_assign(new=c('ph', 'th', 'kh'),
                       copy=c('p_h', 't_h', 'k_h'))
tail(lkup, 3)
```

Say you have been using `ph`, `th`, `kh` for preaspirated stop. They can be 
bulk added like this:

```{r}
lkup <- feature_assign(new=c('hp', 'ht', 'hk'),
                       feature='lar',
                       val='preaspirated',
                       copy=c('p', 't', 'k'))
tail(lkup, 3)
```

Unfortunately, `feature_assign` does not at present allow you to add unknown
IPA characters.

## Contact

If you run into any problems using `phonfeatures`, feel free to leave a bug
report on [GitHub](https://github.com/rpuggaardrode/phonfeatures) or reach out 
at r.puggaard at phonetik.uni-muenchen.de
